---
title: "A Data Visualization Analysis of the Summer Olympic Games"
output:
  html_notebook: default
  pdf_document: default
---

```{r echo=FALSE, message=FALSE, warning=FALSE, fig.align='center'}
library(dplyr)
library(ggplot2)
library(tidyr)
library(forcats)
library(lubridate)
library(sysfonts)
library(showtext)
showtext_auto()
Sys.setlocale("LC_TIME", "C")

font_add_google("PT Sans", "ptsans")
theme_set(theme_minimal())

athlete_events <- read.csv("./data/athlete_events.csv", stringsAsFactors = T)
records_100m <- read.csv("./data/100m_world_records.csv")

```


## Introduction

In this notebook, a selection of exploratory visualizations is presented to investigate key aspects of the Summer Olympic Games.

The dataset used, `athlete_events.csv`, contains detailed information on athletes, disciplines, medals, and participation over time. An additional dataset, `100m_world_records.csv`, tracks the progression of 100m sprint world records since 1977.

The goal is to address several analytical questions through data visualization using `ggplot2` and complementary R packages. Each question is introduced with a specific visualization, followed by an explanation of the plotting technique and the insights derived.

## Datasets

Two datasets are used in this analysis. Both provide complementary views of the Olympic Games: one through detailed athlete records, and the other through the progression of a key athletic performance metric.

**1. Olympic Athlete Events Dataset**

- **Source**: [Kaggle – 120 Years of Olympic History: Athletes and Results](https://www.kaggle.com/datasets/heesoo37/120-years-of-olympic-history-athletes-and-results)  
- **File name**: `athlete_events.csv`  
- **Description**:  
  This dataset contains information on more than 120 years of Olympic history. Each row corresponds to an athlete's participation in an event, including variables such as:
  - Name, sex, age, height, and weight  
  - Team and National Olympic Committee (NOC)  
  - Year, city, and sport  
  - Event and medal received (if any)  
- **Temporal coverage**: 1896–2016  
- **Usage in this project**: Used to analyze athlete achievements, age distributions, and participation trends in the Summer Olympics.

**2. Men's 100m World Record Progression**

- **Source**: [Wikipedia – Men's 100 metres world record progression](https://en.wikipedia.org/wiki/Men%27s_100_metres_world_record_progression)  
- **File name**: `100m_world_records.csv`  
- **Description**:  
  This manually collected dataset includes official world records for the men’s 100m sprint since the introduction of fully electronic timing in 1977. Each entry includes:
  - Record time and date  
  - Athlete name and nationality  
  - Wind conditions  
- **Temporal coverage**: 1977–present  
- **Usage in this project**: Used to explore the evolution of 100 metres world records.

## Question 1: How has the number of athletes evolved over time?

This question explores how the number of athletes participating in the Summer Olympic Games has evolved from the first edition in 1896 to the present day. The aim is to observe long-term trends in athlete participation, with a specific focus on gender differences between male and female athletes.

```{r}

athletes_by_year <- athlete_events %>%
  filter(Season == "Summer") %>%
  select(Year, ID, Sex) %>%
  distinct() %>%
  count(Year, Sex) %>%
  pivot_wider(names_from = Sex, values_from = n, values_fill = 0) %>%
  rename(Men = M, Women = F) %>%
  mutate(Total = Men + Women) %>%
  pivot_longer(cols = c(Men, Women, Total), names_to = "Category", values_to = "Athletes")

custom_colors <- c(
  "Men" = "#EB5E49",       
  "Women" = "#2A9D9C",     
  "Total" = "#555D66"      
)
```


```{r fig.width=12, fig.height=7, echo=TRUE, message=FALSE, warning=FALSE, fig.align='center'}
ggplot(athletes_by_year,
       aes(x = Year, y = Athletes, color = Category))+
  geom_line(size = 1, alpha = 1) +
  geom_point(size = 2) +
  scale_color_manual(values = custom_colors) +
  scale_x_continuous(
    breaks = c(1896,1932,1956,1980,2016)) +
  scale_y_continuous(
    breaks = c(0,1000,5000,10000)) +
  geom_text(aes(label = Athletes ),
            filter(athletes_by_year,Year==2016),
            vjust=-1,
            fontface="bold", show.legend = F)+
  labs(title = "Number of Athletes at the Summer Olympics Over Time",
       subtitle = "Comparison between men, women and total participants",
       x = "Year",
       y = "Number of Athletes",
       color = "Category") +
  annotate("segment", 
       x = 1930, xend = 1932,
       y = max(athletes_by_year$Athletes) * 0.45, 
       yend = athletes_by_year %>% filter(Year == 1932, Category == "Total") %>% pull(Athletes),
       arrow = arrow(length = unit(0.2, "cm")),
       color = "gray50") +
  annotate("text", 
       x = 1952, 
       y = max(athletes_by_year$Athletes) * 0.65, 
       label = "Melbourne: Boycotts for Suez Crisis\nand Soviet invasion of Hungary",
       size = 3.5,
       hjust = 0.5,
       color = "gray20",
       fontface = "italic")+
  annotate("segment", 
       x = 1952, xend = 1956, 
       y = max(athletes_by_year$Athletes) * 0.6, 
       yend = athletes_by_year %>% filter(Year == 1956, Category == "Total") %>% pull(Athletes),
       arrow = arrow(length = unit(0.2, "cm")),
       color = "gray50")+
  annotate("segment", 
       x = 1976, xend = 1980, 
       y = max(athletes_by_year$Athletes) * 0.85, 
       yend = athletes_by_year %>% filter(Year == 1980, Category == "Total") %>% pull(Athletes),
       arrow = arrow(length = unit(0.2, "cm")),
       color = "gray50")+
  annotate("text", 
       x = 1904, 
       y = max(athletes_by_year$Athletes) * 0.35, 
       label = "St. Louis: Low international participation\ndue to location and logistics",
       size = 3.2,
       hjust = 0.5,
       color = "gray20",
       fontface = "italic") +
  annotate("segment", 
       x = 1904, xend = 1904, 
       y = max(athletes_by_year$Athletes) * 0.3, 
       yend = athletes_by_year %>% filter(Year == 1904, Category == "Total") %>% pull(Athletes),
       arrow = arrow(length = unit(0.2, "cm")),
       color = "gray50")+
  annotate("text", 
       x = 1976, 
       y = max(athletes_by_year$Athletes) * 0.90, 
       label = "Moscow: US-led boycott for\nSoviet invasion of Afghanistan",
       size = 3.2,
       hjust = 0.5,
       color = "gray20",
       fontface = "italic")+
  annotate("text", 
       x = 1930, 
       y = max(athletes_by_year$Athletes) * 0.50, ,
       label = "Los Angeles: Great Depression, \ndrop in participation",
       size = 3.5,
       hjust = 0.5,
       color = "gray20",
       fontface = "italic") +
  theme(
    plot.title = element_text(face = "bold", size = 14),
    axis.text.x = element_text(angle = 45, hjust = 1, size = 9),
    legend.position = "top",
    axis.title.x= element_blank(),
    axis.title.y = element_blank(),
    panel.grid.minor.x = element_blank(), 
    panel.grid.minor.y = element_blank(), 
    plot.margin = margin(1, 1, 1, 1, "cm")
  )

sav
```



### Considerations

The chart reveals several key insights:

- There is a consistent upward trend in athlete participation throughout Olympic history.
- Female participation started at zero and gradually increased, reaching nearly the same level as male participation by 2016.
- Certain drops in participation are linked to major geopolitical events, such as the Great Depression, international conflicts, and Olympic boycotts.

### Technical choices

- A line plot was used to display the trend over time, with separate lines for men, women, and the total number of athletes.
- Significant historical events causing notable decreases in participation were annotated directly on the plot to provide additional historical context.


## Question 2: Who are the most decorated athletes in the history of the Summer Olympics?

This question focuses on identifying the top 10 most decorated athletes in the history of the Summer Olympic Games. The objective is to compare athletes based on their total number of medals, as well as the breakdown into gold, silver, and bronze medals.

```{r }


medals_by_athlete <- athlete_events %>%
  filter(Season=="Summer", !is.na(Medal)) %>%
  group_by(ID,Name,NOC, Team, Medal) %>%
  summarise(Count = n(), .groups = "drop") %>%
  pivot_wider(names_from = Medal, values_from = Count, values_fill = 0) %>%
  mutate(Total = (Gold + Bronze + Silver), Label = paste0(Name, " (", NOC, ")")) %>%
  arrange(desc(Total), desc(Gold), desc(Silver), desc(Bronze)) %>%
  slice_head(n = 10)

medals_by_athlete$Label <- medals_by_athlete$Label %>%
  fct_reorder(medals_by_athlete$Bronze) %>%
  fct_reorder(medals_by_athlete$Silver) %>%
  fct_reorder(medals_by_athlete$Gold) %>%
  fct_reorder(medals_by_athlete$Total) 

medals_by_athlete$Label <- recode(medals_by_athlete$Label,
                                  "Larysa Semenivna Latynina (Diriy-) (URS)" = "Larysa Latynina (URS)",
                                  "Jennifer Elisabeth \"Jenny\" Thompson (-Cumpelik) (USA)" = "Jenny Thompson (USA)"
)

plot_data <- medals_by_athlete %>% 
  select(Label, Total, Gold, Silver, Bronze) %>%
  pivot_longer(cols = c(Total, Gold, Silver, Bronze),
               names_to = "Medal",
               values_to = "Count")

plot_data$Medal <- factor(plot_data$Medal,
                          levels = c("Total", "Gold", "Silver", "Bronze"))


medal_colors <- c(
  "Gold"   = "#F4C542",  
  "Silver" = "#A9B8C0",  
  "Bronze" = "#D08C60",  
  "Total"  = "#D6B3FF"   
)

w <- 0.7

```

```{r fig.width=12, fig.height=7, echo=TRUE, message=FALSE, warning=FALSE, fig.align='center'}
ggplot(plot_data, aes(y =Label, x = Count, fill = Medal, colour = Medal)) +
  # Linee (bastoncino del lollipop)
  geom_linerange(aes(xmin = 0, xmax = Count,y=Label),
               position = position_dodge2(width = w,reverse = T),
               size=0.8) +
  geom_point(position = position_dodge2(width = w, reverse = T),
             size=2.5) +
  geom_text(aes(label = Count),
            position = position_dodge2(width = w, reverse = T),
            colour="grey20", fontface="bold", size=3, hjust=-1.3)+
  scale_color_manual(values = medal_colors) +
  scale_fill_manual(values = medal_colors) +
  scale_y_discrete(name="") +
  scale_x_continuous(name="Number of medals", position="top") +
  labs(title="Most decorated athletes in the history of the Summer Olympics") +
  theme_minimal(base_family="ptsans", base_size=10) + 
  theme(panel.grid.minor = element_blank(),
        panel.grid.major.y = element_blank(),
        plot.title = element_text(face="bold",size=12, hjust=0.2),
        axis.text.y = element_text(face="bold")
  )
```




### Considerations

The chart highlights several important insights:
- Michael Phelps (USA) stands out as the most decorated athlete, with 28 medals, including a record 23 golds.
- Athletes from the former Soviet Union (URS), shows historical dominance.


### Technical choices

- A lollipop plot was selected to clearly display the count of medals for each athlete, separated by medal type (Gold, Silver, Bronze) and total.
- The athletes are sorted from top to bottom based on the total number of medals. In case of ties, a hierarchical sorting is applied: first by the number of gold medals, then silver, and finally bronze.
- The number of medals is displayed next to each lollipop marker to improve clarity
- The National Olympic Committee (NOC) code is displayed next to each athlete's name

## Question 3: What is the distribution of athlete ages by sport at the Rio 2016 Olympics?

This question investigates how the ages of athletes varied across different sports during the 2016 Summer Olympic Games in Rio de Janeiro. 

The aim is to identify age-related patterns within each discipline and highlight which sports tend to involve younger or older participants.

```{r}

athlete_events <- read.csv("./data/athlete_events.csv", stringsAsFactors = FALSE)
breaks <- seq(10, 64, by = 4)

labels <- paste0("[", breaks[-length(breaks)], "–", breaks[-1] - 1, "]")

heatmap_data <- athlete_events %>%
  filter(Year == 2016, Season == "Summer", !is.na(Age))


heatmap_data <- heatmap_data %>%
  mutate(
    Age_Group = cut(Age,
                    breaks = breaks,
                    labels=labels,
                    include.lowest = TRUE,
                    right = FALSE)
  ) %>%
  group_by(Sport) %>%
  mutate(Total_Sport = n()) %>%
  group_by(Sport, Age_Group) %>%
  summarise(
    Participants = n(),
    Total_Sport = first(Total_Sport),
    .groups = "drop"
  ) %>%
  complete(Sport, Age_Group, fill = list(Participants = 0)) %>%
  group_by(Sport) %>%
  mutate(
    Total_Sport = ifelse(is.na(Total_Sport), sum(Participants), Total_Sport),
    Percentage = ifelse(Total_Sport == 0, 0, (Participants / Total_Sport) * 100)
  ) %>%
  ungroup()



```

```{r fig.width=9, fig.height=6, echo=TRUE, message=FALSE, warning=FALSE, fig.align='center'}
ggplot(heatmap_data, aes(x = Age_Group, y = fct_rev(Sport), fill = Percentage)) +
  geom_tile() +
  scale_fill_viridis_c(
    option = "viridis",      
    trans = "sqrt",          
    limits = c(0, max(heatmap_data$Percentage, na.rm = TRUE)),
    name = "% of Athletes"
  ) +
  labs(
    title = "Distribution of Athlete Ages by Sport (Rio 2016)",
    subtitle = "Percentage of participants per age group within each sport",
    x = "Age Group",
    y = "Sport"
  ) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    panel.grid = element_blank(),
    axis.title.x = element_blank(),
    axis.title.y = element_blank()
  )
```

### Considerations

The chart provides several noteworthy insights:

- In most sports, the majority of athletes fall within the 22–29 age group, reflecting the general peak of physical performance.
- Some sports, such as Shooting and Equestrianism, show a broader age distribution and include participants even in the 58–61 age range.
- Rhythmic Gymnastics stands out as the sport with the most homogeneous age distribution, with over 60% of its athletes aged between 18 and 21.

### Technical choices

- A heatmap was used to represent the percentage of athletes per age group within each sport.
- Age ranges are defined in 4-year intervals, and percentages are calculated within each sport to allow for comparability.

## Question 4: How has the 100m world record evolved over time?

This question examines the progression of the men's 100m world record from the beginning of fully electronic timing in 1968 to the present. The objective is to understand the evolution of sprint performance over time and to identify patterns in athlete nationalities and performance improvements.

```{r}
records <- records_100m %>%
  mutate(Date = mdy(Data)) %>%
  arrange(Date) %>%
  mutate(End = lead(Date)) %>%
  mutate(End = if_else(is.na(End), Sys.Date(), End)) %>%
  select(Time, Wind, Athlete, Nationality, Date, End)


min_date <- min(records$Date)
max_date <- max(records$End)

```

```{r fig.width=16, fig.height=7, echo=TRUE, message=FALSE, warning=FALSE, fig.align='center'}
ggplot(records, aes(y = Time, colour = Nationality)) + 
  geom_point(aes(x = Date), size=1.5) +
  geom_point(data = records %>% filter(End != Sys.Date()), aes(x = End), size=1.5) +
  geom_text(aes(label = Athlete, x = End),data=records %>% filter(End != Sys.Date()), hjust = -0.1, angle = 0, show.legend = F) +
  geom_text(aes(label = Athlete, x = End),data=records %>% filter(End == Sys.Date()),fontface="bold", hjust = 1,vjust=1.4, angle = 0, show.legend = F) +
  geom_text(aes(label = format(Date, "%d %B %Y"), x = Date), 
            colour = "grey20", hjust = 1.2, angle = 0, show.legend = F) +
  geom_text(data = records %>% filter(End == Sys.Date()),
              aes(x = End, y = Time, label = "Current World Record"),
              hjust = 1, vjust = -0.7, colour = "grey20", fontface = "bold", size = 4, show.legend = F)+
  geom_segment(aes(x = Date, xend = End, y = Time, yend = Time), size=0.6) +
  geom_line(aes(x = Date, y = Time), size = 2, alpha = 0.3, colour = "gray80") + 
  scale_y_continuous(breaks = sort(unique(records$Time))) +
  scale_x_date(limits = c(min_date - years(3), max_date)) +
  scale_color_manual(values = c(
    "United States" = "#1f77b4",
    "Jamaica"       = "#1cae49",
    "Canada"        = "#e63946"
  )) +
  labs(
    title = "Official 100m World Record Progression (4 October 1968– 04 June 2025, Electronic Timing Only)",
    subtitle = "Since 1977, only fully electronic times measured to the hundredth of a second are recognized as official world records.",
    x = "Date",
    y = "Time (s)",
  )+
  theme_minimal(base_family = "ptsans") + 
  theme(axis.title.x = element_blank(),
        panel.grid.minor = element_blank(),
        panel.grid.major.x = element_blank(),
        axis.text.x = element_blank(),
        plot.title = element_text(face="bold",size=12, hjust=0.2),
        plot.caption = element_text(size=10),
        axis.text.y = element_text(face="bold", size=10))
```

### Considerations

The chart reveals several important observations:
- Since the adoption of fully electronic timing, only three countries have held the 100m world record: the United States, Canada, and Jamaica.
- U.S. athletes dominated the record progression for many years, followed by a brief period in which a Canadian sprinter held the record, and more recently, Jamaican athletes have taken over.
- From 1968 to 2009, the improvement in the world record has been less than four-tenths of a second, highlighting the difficulty in achieving further reductions in time at the elite level.

### Technical choices

- A line plot was used to illustrate the progression of the record times over the years
- Colors were used to indicate the nationality of each record-holding athlete.
- Dumbbell-style segments were added to emphasize the duration each record remained unbroken. The last record by Usain Bolt does not include an endpoint marker, visually reinforcing that it is the current record.




